¥section{Methodology}
¥label{sec:method}

¥subsection{Models}
¥label{sec:method:models}

We compare three models, summarized in ¥tabref{tab:models}. ¥talkienineteen and ¥talkieweb are matched-architecture, matched-FLOPs 13B decoder-only transformers (40 layers, 40 heads, embedding dimension 5120, vocabulary size 65{,}536, RoPE, SwiGLU, RMSNorm) released by ¥citet{talkielm2026}. The only intended difference is the pretraining corpus: pre-1931 English text (260B tokens, Common Pile + IDI + Internet Archive) for ¥talkienineteen, and modern FineWeb for ¥talkieweb. We load each in bf16 on a single NVIDIA RTX A6000 (48‾GB VRAM) and decode greedily; we bypass the upstream sampler that injects Gumbel noise even at temperature zero. ¥gptfour is queried via the OpenAI API at ¥texttt{temperature=0}.

¥begin{table}[h]
    ¥centering
    ¥caption{Models compared. Both Talkie checkpoints share architecture and compute; only the pretraining corpus differs.}
    ¥label{tab:models}
    ¥begin{tabular}{@{}lcccc@{}}
        ¥toprule
        ¥textbf{Model} & ¥textbf{Size} & ¥textbf{Cutoff} & ¥textbf{Corpus} & ¥textbf{Access} ¥¥
        ¥midrule
        ¥talkienineteen & 13B & 1930 & 260B tokens pre-1931 English & HF weights ¥¥
        ¥talkieweb & 13B & modern & FineWeb (matched arch + FLOPs) & HF weights ¥¥
        ¥gptfour & --- & unknown & unknown & OpenAI API ¥¥
        ¥bottomrule
    ¥end{tabular}
¥end{table}

¥subsection{Datasets}
¥label{sec:method:datasets}

¥para{Post-1930 probe battery.} We construct 50 items: 30 post-1930 factual completions (¥eg ``On August 6, 1945, the United States dropped¥dots''), 9 pre-1930 controls (¥eg ``The American Civil War began in the year¥dots''), and 11 in-context-learning items in Python and JavaScript (each prompt contains one to three worked examples followed by a continuation cue). All items are written by hand or constructed by a small script and verified manually. The probe battery is released as part of our reproduction pipeline.

¥para{Date-stratified ¥mmlu.} We subsample 456 ¥mmlu items, 8 per subject across all 57 subjects, with a fixed seed. Each item is auto-tagged ¥texttt{post1930}, ¥texttt{pre1930}, or ¥texttt{unknown} by a conservative regex+keyword detector applied to the question text only (not answer choices, to avoid numeric-choice false positives). Year tokens use a 4-digit pattern with negative look-around to skip decimals; a hand-curated keyword whitelist captures unambiguously post-1930 vocabulary (¥texttt{internet}, ¥texttt{wifi}, ¥texttt{CRISPR}, ¥texttt{polio vaccine}, ¥dots), with an upper-case-only acronym list (¥texttt{DNA}, ¥texttt{NATO}, ¥dots) to avoid matching lowercase homonyms. On the full ¥mmlu test set, the heuristic tags 12.7¥% post-1930, 2.6¥% pre-1930, and 84.7¥% unknown.

¥para{¥humaneval.} 164 Python function-completion problems with hidden unit tests‾¥citep{chen2021humaneval}. We evaluate pass@1 with $k ¥in ¥{0, 3¥}$ in-context demonstrations.

¥subsection{Scoring}
¥label{sec:method:scoring}

¥para{Probes.} For each probe we compute (i) the per-token bits-per-token of the expected modern continuation in one forward pass over (prompt + target), and (ii) the greedy-decoded continuation (max 32 tokens, stop on newline). The default ¥texttt{TalkieModel.forward} returns only the last-position logits; we wrote ¥texttt{¥_forward¥_all¥_logits} to recover all-position logits in a single pass. For ¥gptfour we use the Chat Completions API and check for the expected modern token under a ¥emph{lenient} exact-match rule (case-fold, punctuation-strip, substring).

¥para{¥mmlu.} For Talkie models, we score the four single-token candidates $¥{$``¥,A''$, $``¥,B''$, $``¥,C''$, $``¥,D''$¥}$ from the next-token distribution at the end of a standard ``Question / A / B / C / D / Answer:'' prompt (one forward pass per item). For ¥gptfour we prompt for a letter and parse it.

¥para{¥humaneval.} Greedy decoding with ¥texttt{max¥_tokens=192}, stop-strings ¥{¥texttt{$¥backslash$ndef}, ¥texttt{$¥backslash$nclass}, ¥texttt{$¥backslash$nif ¥_¥_name¥_¥_}, ¥texttt{$¥backslash$nimport}, ¥texttt{$¥backslash$nfrom}, ¥texttt{$¥backslash$n¥#}, ¥texttt{$¥backslash$n$¥backslash$n$¥backslash$n}¥}, plus a special prefix-stop rule to handle the case where the bare base model immediately emits ¥texttt{def foo():} (re-defining the prompt's function instead of writing its body --- observed empirically). Generated code is evaluated by the official ¥texttt{check(candidate)} function in a sandboxed subprocess.

¥subsection{Statistical Tests}
¥label{sec:method:stats}

For every proportion estimate (pass@1, EM, accuracy) we report 95¥% bootstrap confidence intervals (2{,}000 resamples). For paired per-item ¥bpt comparisons between ¥talkienineteen and ¥talkieweb we use a paired permutation test (5{,}000 sign-flip permutations, two-sided). All tests use seed 0.

¥subsection{Attribution-Bits Framework}
¥label{sec:method:bits}

The load-bearing definitional move of the meta-experiment is to convert the structural cutoff argument into a quantitative attribution score. For a single correctly answered probe item, the Bayes factor in favor of ``the model generalized'' over ``the model memorized'' is
¥begin{equation}
¥mathrm{BF}_{¥text{gen}} ¥;=¥; ¥frac{P(¥text{correct} ¥mid ¥text{gen}) ¥cdot P(¥text{gen})}{P(¥text{correct} ¥mid ¥text{memo}) ¥cdot P(¥text{memo})}.
¥label{eq:bf}
¥end{equation}
We set $P(¥text{correct} ¥mid ¥text{memo}) = 1$ (a perfect memorizer trivially recovers anything in its training data), $P(¥text{correct} ¥mid ¥text{gen}) ¥approx p_{¥text{correct}}$ (the model's empirical accuracy on the probe family), and equal priors $P(¥text{gen}) = P(¥text{memo})$. The remaining free parameter is the prior probability $P(¥text{memo})$ that the item is in training.

¥para{Why ¥talkienineteen's structural cutoff is load-bearing.}
For ¥talkienineteen on a post-1930 item, $P(¥text{memo}) ¥approx 0$ by construction. We cap at $P(¥text{memo}) = 10^{-12}$ for finite arithmetic, yielding $¥log_2 ¥mathrm{BF}_{¥text{gen}} ¥approx 34$ bits of evidence per correct answer. For ¥talkieweb and ¥gptfour on the same item, the training corpus is unknown and we must use $P(¥text{memo}) = 1$ as the worst-case value: bits become $¥le 0$, i.e., the observation provides no evidence against memorization. The asymmetry between the two arms is the meta-experimental claim: ¥talkienineteen trades a hard structural assumption for the freedom to compute the bits at all. Modern-LLM setups must spend additional resources --- n-gram indices, contamination audits --- to lower $P(¥text{memo})$ before they can produce a non-trivial number.
